---
params:
  directory: ''
  sampleName: ''
output:
  officedown::rdocx_document:
    reference_docx: ../docs/MABA16S_template.docx
---

```{r packages, include=FALSE}

library(tidyverse)
library(flextable)
library(ggplot2)
library(openxlsx)
library(stringr)
#library(data.table)

```

```{r flextable, include = FALSE}

set_flextable_defaults(
  font.family = "calibri",
  font.size = 10,
  font.color = "black",
  border.color = "grey20",
  border.width = 0.5,
  padding.top = 3,
  padding.bottom = 3,
  padding.left = 5,
  padding.right = 5,
  line_spacing = 1,
  background.color = "transparent",
  table.layout = "autofit",
  digits = 2,
  na_str = "-"
)

```

```{r Functions, include = FALSE}

# function to look for alternative blast hits if the first hit is uncultured bacterium
second_or_first_key <- function(x) {
  # Extract keys that look like: 'some text':
  keys <- str_match_all(x, "'([^']+)'\\s*:")[[1]][, 2]

  if (length(keys) >= 2) {keys[2]
    } else if (length(keys) == 1) {keys[1]
  } else {
    NA_character_}}

```

```{r setPATHS, include = FALSE}

# set data directory
dir = params$directory

# set sample name
sample = params$sampleName

```

```{r sampleSheet, include = FALSE}

# retrieve original sample barcode from the name conversion sample sheet
samplesheet = data.table::fread(paste0(dir, "/MABA16S_sampleSheet.tsv"))
  
# Retrieve all NTCs from the sample sheet
ntcs = samplesheet %>% filter(grepl("NTC", SampleName, ignore.case = T))

# set the order in which the samples should be displayed
order = c(sample, ntcs$SampleName)

```

```{r preprocessingQC, include = FALSE}

# retrieve the generated QC output from the pre-analysis pipeline
firstQC = do.call(rbind, 
                    lapply(c(sample, ntcs$SampleName), 
                            function(s) data.table::fread(paste0(dir, "/QC/", s, "/", s, "_qcPreprocessing.txt"))))

# This also contains the information on how many kraken entries generated at least
# 50 reads and were therefore analysed further

# Add an explainer column for later inclusion in the report
firstQC = firstQC %>% mutate(Note = ifelse(nGenera == 0,
                                           "No genera analysed", "-"))

# Determine if the sample had enough kraken hits to be fully analysed
analysed = firstQC[sampleID == sample, nGenera] > 0

# Determine which samples were analysed (at least 1 qualifying genus)
inclusion = firstQC %>% filter(nGenera > 0)

# check if any NTCs have failed
ntcExcl = ntcs %>% filter(!SampleName %in% inclusion$sampleID)

```

```{r postprocessingQC, include = FALSE}

# This is needed to check that all genera were analysed

# retrieve the QC output from the analysis pipeline
finalQC = do.call(rbind, 
                    lapply(c(inclusion$sampleID), 
                            function(s) data.table::fread(paste0(dir, "/QC/", s, "/", s, "_qcPostAnalysis.txt"))))

##### need to remove extra header line from the postAnalysisQC

# retain only entries where not all genera were successfully analysed
finalQC = finalQC %>%
  mutate(
    allAnalysed =
      nGenera == nGenusReads &
      nGenera == nAligned &
      nGenera == nConsensusFastas &
      nGenera == nBlast
  ) %>%
  filter(allAnalysed == FALSE)

# Make a variable for whether all genera were correctly analysed
generaCheck = nrow(finalQC) < 1

```

```{r continue, include = F}

# Check if the conditions to continue to include the analyses results were met
continue = analysed == TRUE & generaCheck == TRUE

```

```{r fetchHits, eval = continue, include = FALSE}

# This chunk only runs if the sample was fully analysed
# Only NTCs with at least 1 analysed genus are included

# read and combine the data from the individual reports 
allData = do.call(rbind,
                  lapply(inclusion$sampleID,
                         function(s) read.xlsx(paste0(dir, "/reports/", s, ".xlsx"),
                                               na.strings = "") %>%
                           mutate(SampleName = s)))

# process the hits
allData = allData %>%
  mutate(
    # Differentiate between spike-ins & results
    spikeIn = factor(ifelse(grepl("imtechella|truepera|allobacillus", blast_hit, ignore.case = TRUE), 
                            "Spike-in", "Other"),
                     levels = c("Spike-in", "Other")),
    
    # Add renaming logic
    FinalName = case_when(
      # ensure that the spike-in reads are appropriately grouped
      grepl("imtechella", blast_hit, ignore.case = TRUE) ~ "Imtechella",
      grepl("trueper", blast_hit, ignore.case = TRUE) ~ "Truepera",
      grepl("allobacillus", blast_hit, ignore.case = TRUE) ~ "Allobacillus",
      
      # try to rename uncultured bacteria if alternative options available
      blast_hit == "uncultured bacterium" ~ second_or_first_key(blast_hits),
      TRUE ~ blast_hit
      ),
    
    # Neaten up blast_hits string for inclusion in the report
    blast_hits = blast_hits %>%
      str_remove("^\\{") %>%
      str_remove("\\}$")
  ) %>%
  # calculate a cumulative count
  group_by(SampleName, FinalName) %>%
  mutate(speciesReads = sum(num_reads)) %>%
      ungroup() %>%
  as.data.frame()

```

# MUMC+ MMI: 16S ONT

## Single sample report

__Date:__ `r format(Sys.Date(), '%d-%m-%Y')`

__Sample ID:__ `r sample`

### Version information:

__Maastricht bacterial 16S ONT pipeline version:__ 1.2.0

### Method:

The sample underwent DNA extraction, 16S amplification and sequencing on the Oxford Nanopore GridIon.  

A 10.4.1 flow cell and chemistry was used.  

__Flow cell type:__ placeholder & __flow cell ID:__ placeholder  

__Kit type:__ placeholder & __Kit lot number:__ placeholder  

__ONT sequencing barcode - sample:__ `r samplesheet$Barcode[samplesheet$SampleName == sample]`  

__Negative controls:__ `r ntcs$SampleName` from barcodes `r ntcs$Barcode` respectively.  

### Conclusion:
###### <<space for text, please replace>>

### Results overview

`r if (!analysed) "###### This sample generated no kraken2 genus hits with a minimum of 50 reads and was therefore not analysed further"`

`r if (nrow(ntcExcl) > 0)
  paste0(
    "###### NTC ",
    ntcExcl$SampleName,
    " is not shown because it generated no kraken2 genus hits with a minimum of 50 reads and was therefore not analysed further"
  )`

`r if (!generaCheck) "###### At least one genus failed to analyse correctly. Please ask the bioinformatician to check the analysis"`

```{r resultsVisualisation, eval = continue, echo = F}

plot = allData %>% 
  ggplot(aes(x = reorder(blast_hit, -speciesReads), y = num_reads, fill = spikeIn)) +
      geom_col(position = "stack", colour = "black") +
      geom_vline(xintercept = -Inf) +
      geom_hline(yintercept = -Inf) +
      facet_grid(SampleName ~ spikeIn, scales = "free_x", space = "free_x") +
      theme_classic() +
      scale_fill_manual(values = c("Spike-in" = "grey70", "Other" = "chartreuse3"), 
                        guide = "none") +
      scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
      scale_x_discrete(expand = c(0,0)) +
      ylab("number of reads") +
      theme(axis.title.x = element_blank(),
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
            axis.text.y = element_text(size = 8),
            axis.title.y = element_text(size = 9),
            axis.line = element_line(),
            strip.text = element_text(size = 9, face = "bold"),
            strip.background = element_rect(colour = "grey70", fill = "grey90"))
    
    # save the plot
    ggsave(plot, filename = paste0(dir, "/barCharts/", sample, "_CompositionBarChart.png"),
           width = (length(unique(allData$blast_hit)) * 0.4) + 1.5, height = 2.2 + (length(unique(allData$SampleName))* 2.2), dpi = 300)
    
    # add the plot to the report
    plot

```

### Detailed summary & QC information

```{r qcTable, echo = F}

# Format and print the QC table
firstQC %>% 
  select(-nGenera) %>%
  rename("Sample Name" = sampleID,
        "Raw reads" = rawReads,
        "Filtered reads" = filteredReads) %>%
  flextable() %>%
  #format the header & body
  bold(part = "header") %>%
  fontsize(part = "header", size = 11) %>%
  align(align = "center", part = "all", j = c(1:4)) %>%
  #add conditional formatting
  bg(i = ~ Note == "No genera analysed", j = c("Sample Name", "Note"), bg = "firebrick1") 

```

## End of report
